Analysis of NLA17 Dissolved Gas Indicator

### 2019-11-15 scripts/dgIndicatorAnalysis.Rmd
Added population estimate figures and variograms to .Rmd

---
title: "Analysis of NLA17 Dissolved Gas Indicator"
author: "J. Beaulieu"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    toc: true
    toc_float: false
    fig_caption: yes
    depth: 2
    number_sections: true
---

```{r results='hide', message=FALSE, warning=FALSE}
# load libraries
library(rgdal) # spatial data
library(sf) # spatial data
library(gstat) # lagged scatterplot
library(spdplyr) # apply dplyr verbs to sp objects
library(rmarkdown)
library(tidyverse)
```

@@

The dissolved gas data were currated the 'NLA' dissolved gas project in RStudio.  The code can be found at a private github repository (https://github.com/USEPA/NLA).  After aggregating across duplicate samples, the data were written to nla17gasDataAggregated_2019-10-25.txt.  A copy of this data file is stored in documents library associated with the 'ORD NLA17 Dissolved Gas' Private Group on SharePoint (https://usepa.sharepoint.com/sites/ORD_NLA17_Dissolved_Gas).  When synced to a local computer, the documents library can be read directly with R, after identifying the local directory path.

```{r}
# Identify local path for each user
localPath <- Sys.getenv("USERPROFILE")

# Read data file
dg <- read.table(file = paste0(localPath, 
                               "/Environmental Protection Agency (EPA)/",
                               "ORD NLA17 Dissolved Gas - Documents/",
                               "inputData/nla17gasDataAggregated_2019-10-25.txt"), 
                 header = TRUE, sep = "\t", as.is = TRUE)
```
Concentrations of CO~2~, CH~4~, and N~2~O in ambient air over the waterbodies were measured and used in the dissolved gas concentrations.  These data are not needed in subsequent analysis and will be removed from the dataframe.
```{r}
dg <- dg %>% filter(sample.source == "DG") # filter out dissolved gas 'DG' samples
```
The dataframe also contains observations from site in Alaska.  These sites were not an official part of the NLA and will not be included in this analysis.
```{r}
dg <- dg %>% filter(!(grepl("AK", site.id))) # exlclude alaska
```

The data file contains `r dg %>% distinct(site.id, visit.no) %>% summarize(n = n()) %>% pull()` unique sampling trips defined by unique combinations of 'site.id' and 'visit.no'.  There were `r dg %>% distinct(site.id) %>% summarize(n = n()) %>% pull()` waterbodies sampled, `r dg %>% filter(visit.no == 2) %>% summarize(n = n()) %>% pull()` of which were sampled twice. The primary response variables are the concentrations of CO~2~, CH~4~, and N~2~O dissolved in the surface water of each waterbody.  Concentration is expressed as both mol/L (dissolved.co2, dissolved.ch4, dissolved.n2o) and as a saturation ratio (co2.sat.ratio, ch4.sat.ratio, n2o.sat.ratio).  Saturation ratio is defined as the ratio of the measured concentration to that expected if the waterbody were in equilibrium with the atmosphere.  A value of 1 indicates equilibrium, values < 1 indicate undersaturation and values > 1 indicate supersaturation.  Undersaturated waterbodies function as a sink for the atmospheric gas whereas supersaturated waterbodies function as a source.

@@

To enable spatial analysis of the data, the dataframe will be converted to a 'simple features' (sf) object.
```{r}
# Define coordinates
coords <- data.frame(longitude = dg$map.lon.dd, latitude = dg$map.lat.dd)

dg.sf <- st_as_sf(dg, coords = c("map.lon.dd", "map.lat.dd"), 
                  crs = 4269) %>% # standard for lat/lon
  st_transform(5070) # project to CONUS ALBERS for plotting
```

#  Exploratory spatial analysis

Bubble plots offer a simple method for visualizing spatial patterns in the data.  For some reason, this isn't working.  WTF....

```{r}
# read in ecoregion polygons
ecoR <- st_read(dsn = paste0(localPath, 
                             "/Environmental Protection Agency (EPA)/",
                             "ORD NLA17 Dissolved Gas - Documents/inputData"),
                layer = "Aggr_Ecoregions9_2015")

@@

# Check CRS
st_crs(ecoR) # no EPSG code
ecoR <- st_transform(ecoR, 5070) # convert to CONUS Albers
st_crs(ecoR) # 5070

# Bubble plots
ggplot() +
  geom_sf(data = ecoR, color = NA, aes(fill = WSA9_NAME)) +
  geom_sf(data = dg.sf, aes(size = dissolved.co2)) +
  ggtitle("Dissolved CO2")

ggplot() +
  geom_sf(data = ecoR, color = NA, aes(fill = WSA9_NAME)) +
  geom_sf(data = dg.sf, aes(size = dissolved.ch4)) +
  ggtitle("Dissolved CH4")

ggplot() +
  geom_sf(data = ecoR, color = NA, aes(fill = WSA9_NAME)) +
  geom_sf(data = dg.sf, aes(size = dissolved.n2o)) +
  ggtitle("Dissolved N2O")
```

>>>>

---
title: "Analysis of NLA17 Dissolved Gas Indicator"
author: "J. Beaulieu"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    toc: true
    toc_float: true
    fig_caption: yes
    depth: 2
    number_sections: true
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

```{r results='hide', message=FALSE, warning=FALSE}
# load libraries
library(sf) # spatial data
library(gstat) # lagged scatterplot, variogram
#library(spdplyr) # apply dplyr verbs to sp objects.  gstat functions require sp objects
library(tidyverse) # dplyr, ggplot

# Identify local path for each user
localPath <- Sys.getenv("USERPROFILE")

# Define helper functions
# standardized formatting for column names
toEPA <- function(X1){
  names(X1) = tolower(names(X1))
  names(X1) = gsub(pattern = c("\\(| |#|)|/|-|\\+|:|_"), replacement = ".", x = names(X1))
  X1
}
```

@@

The dissolved gas data were currated in the 'NLA' dissolved gas project in RStudio.  The code can be found at a private github repository (https://github.com/USEPA/NLA).  After aggregating across duplicate samples, the data were written to nla17gasDataAggregated_2019-10-25.txt.  A copy of this data file is stored in the documents library associated with the 'ORD NLA17 Dissolved Gas' Private Group on SharePoint (https://usepa.sharepoint.com/sites/ORD_NLA17_Dissolved_Gas).  When synced to a local computer, the documents library can be read directly with R.

```{r}
# Read data file
dg <- read.table(file = paste0(localPath, 
                               "/Environmental Protection Agency (EPA)/",
                               "ORD NLA17 Dissolved Gas - Documents/",
                               "inputData/nla17gasDataAggregated_2019-10-25.txt"), 
                 header = TRUE, sep = "\t", as.is = TRUE) %>%
  # unit conversion
  mutate(dissolved.ch4 = dissolved.ch4 * 1000000, # mol/L -> umol/L
         dissolved.co2 = dissolved.co2 * 1000000, # mol/L -> umol/L,
         dissolved.n20 = dissolved.n2o * 1000000000, # mol/L -> nmol/L
         # add source/sink column
         co2.src.snk = ifelse(co2.sat.ratio > 1, "source", "sink"),
         ch4.src.snk = ifelse(ch4.sat.ratio > 1, "source", "sink"),
         n2o.src.snk = ifelse(n2o.sat.ratio > 1, "source", "sink"))
```
Concentrations of CO~2~, CH~4~, and N~2~O in ambient air over the waterbodies were measured and used in the dissolved gas concentrations.  These data are not needed in subsequent analysis and will be removed from the dataframe.
```{r}
dg <- dg %>% filter(sample.source == "DG") # select dissolved gas 'DG' samples
```
The dataframe also contains observations from site in Alaska.  These sites were not an official part of the NLA and will not be included in this analysis.
```{r}
dg <- dg %>% filter(!(grepl("AK", site.id))) # exlclude alaska
```

The data file contains `r dg %>% distinct(site.id, visit.no) %>% summarize(n = n()) %>% pull()` unique sampling trips defined by unique combinations of 'site.id' and 'visit.no'.  There were `r dg %>% distinct(site.id) %>% summarize(n = n()) %>% pull()` waterbodies sampled, `r dg %>% filter(visit.no == 2) %>% summarize(n = n()) %>% pull()` of which were sampled twice. The primary response variables are the concentrations of CO~2~, CH~4~, and N~2~O dissolved in the surface water of each waterbody.  Concentration is expressed in both mass based units (umol/L, nmol/L: dissolved.co2, dissolved.ch4, dissolved.n2o) and as a saturation ratio (co2.sat.ratio, ch4.sat.ratio, n2o.sat.ratio).  Saturation ratio is defined as the ratio of the measured concentration to that expected if the waterbody were in equilibrium with the atmosphere.  A value of 1 indicates equilibrium, values < 1 indicate undersaturation and values > 1 indicate supersaturation.  Undersaturated waterbodies function as a sink for the atmospheric gas whereas supersaturated waterbodies function as a source.

@@

#  Exploratory spatial analysis
## Population estimates by ecoregion
The survey design allows for population estimates of central tendency and variance by ecoregion (and other subpopulations), which can provide insight into spatial patterning.  There is broad overlap in mean +/- 95% CI estimates for dissolved CH4 concentation.  The Northern Plains, Temperate Plains, and Western Mountains have low dissolved CO2 concentration.

```{r, fig.cap= "Fig. caption: Mean +/- 95% CI for the dissolved gas concentration indicator"}
# Read population estimates provided by Steve Paulsen (11/7/2019)
pop <- read.csv(file = paste0(localPath, 
                              "/Environmental Protection Agency (EPA)/",
                              "ORD NLA17 Dissolved Gas - Documents/",
                              "inputData/populationEstimates/",
                              "NLA_2017_Percentile_Estimates_20191107.csv"),
                as.is = TRUE) %>%
  toEPA()  # enforce formatting convention

# Convert units for dissolved gas concentration indicator
pop <- pop %>%
  mutate(estimate = ifelse(grepl("dissolved.c", indicator, ignore.case = TRUE), # dissolved.co2 or dissolved.ch4
                            estimate * 1000000, # mol/L -> umol/L for co2 and ch4
                           ifelse(grepl("dissolved.n", indicator, ignore.case = TRUE), # dissolved.n2o
                            estimate * 1000000000, # mol/L -> nmol/L for n2o
                            estimate)), # all others
         lcb95pct = ifelse(grepl("dissolved.c", indicator, ignore.case = TRUE), # dissolved.co2 or dissolved.ch4
                            lcb95pct * 1000000, # mol/L -> umol/L for co2 and ch4
                           ifelse(grepl("dissolved.n", indicator, ignore.case = TRUE), # dissolved.n2o
                            lcb95pct * 1000000000, # mol/L -> nmol/L for n2o
                            lcb95pct)), # all others
         ucb95pct = ifelse(grepl("dissolved.c", indicator, ignore.case = TRUE), # dissolved.co2 or dissolved.ch4
                            ucb95pct * 1000000, # mol/L -> umol/L for co2 and ch4
                           ifelse(grepl("dissolved.n", indicator, ignore.case = TRUE), # dissolved.n2o
                            ucb95pct * 1000000000, # mol/L -> nmol/L for n2o
                            ucb95pct))) # all others

# extract population estimates for ecoregion and nation
pop.eco <- pop %>% 
  filter(type %in% c("AG_ECO9_NM", "National"), # extract estimates for ecoregion and nation
         statistic == "50Pct") # extract mean estimate

# extract dissolved gas indicator, order by dissolved CH4
pop.eco.diss <- pop.eco %>%
  filter(grepl("dissolved", indicator, ignore.case = TRUE)) %>%
  mutate(subpopulation = factor(subpopulation, 
                             levels = filter(pop.eco, indicator == "DISSOLVED_CH4", subpopulation != "National") %>%
                               arrange(estimate) %>%
                               select(subpopulation) %>%
                               pull() %>%
                               c("National", .)))

# look at dissolved concentration indicator
ggplot(pop.eco.diss, aes(estimate, subpopulation)) +
  geom_point() +
  geom_errorbarh(aes(xmax = ucb95pct, xmin = lcb95pct)) +
  facet_wrap(~indicator, scales = "free_x") +
  theme(axis.title = element_blank())

```

## Bubble plots
Bubble plots can be used to visualize patterns in geospatial data.  In the plots below, bubble size reflects the gas concentration and color reflects saturation status (i.e. source or sink).  The code immediately below converts the dataframe to a spatial object and reads in the ecoregion ploygons.  The first bubble plot below is of dissolved CO2 which seems to show a pattern of higher concentrations along the eastern seaboard.

```{r}
# To enable spatial analysis of the data, the dataframe will be converted to a 'simple features' (sf) object.
# Define coordinates
coords <- data.frame(longitude = dg$map.lon.dd, latitude = dg$map.lat.dd)

dg.sf <- st_as_sf(dg, coords = c("map.lon.dd", "map.lat.dd"), 
                  crs = 4269) %>% # standard for lat/lon
  st_transform(5070) # project to CONUS ALBERS for plotting

# read in ecoregion polygons
ecoR <- st_read(dsn = paste0(localPath, 
                             "/Environmental Protection Agency (EPA)/",
                             "ORD NLA17 Dissolved Gas - Documents/inputData"),
                layer = "Aggr_Ecoregions9_2015")

@@

# Check CRS
st_crs(ecoR) # no EPSG code
ecoR <- st_transform(ecoR, 5070) # convert to CONUS Albers
st_crs(ecoR) # 5070
```

```{r}
# CO2 Bubble plots
ggplot() +
  geom_sf(data = ecoR, color = NA, aes(fill = WSA9_NAME)) +
  geom_sf(data = dg.sf, aes(size = dissolved.co2, color = co2.src.snk), 
          show.legend = "point") + # improve legend
  ggtitle("Dissolved CO2 (uM)") +
  scale_color_manual(values = c("white", "black"), name = "source/sink") +
  scale_size(name = "dissolved CO2 (uM)",
             range = c(0.1, 5), # custom size range
             breaks = c(10, 100, 300, 600)) # custom breaks
```

Spatial patterns in dissolved CH4 are not apparent, although all but two sites are supersaturated.

```{r}
ggplot() +
  geom_sf(data = ecoR, color = NA, aes(fill = WSA9_NAME)) +
  geom_sf(data = dg.sf, aes(size = dissolved.ch4, color = ch4.src.snk), 
          show.legend = "point") +
  ggtitle("Dissolved CH4 (uM)") +
  scale_color_manual(values = c("white", "black"), name = "source/sink") +
  scale_size(name = "dissolved CH4 (uM)",
             range = c(0.1, 10), # custom size range
             breaks = c(0.1, 1, 10, 50, 100)) # custom breaks
```

The prevalance of undersaturated dissolved N2O concentrations is an important finding.  The IPCC assumes surface waters are an important source of N2O to the atmosphere.   Webb et al. recently published a paper in PNAS reporting that 70 out of 101 reservoirs sampled in Saskatchewan were N2O sinks.  This paper made quite a splash.  In our study, `r sum(dg$n2o.src.snk == "sink")` out of `r sum(!is.na(dg$dissolved.n2o))` sampled waterbodies were an N2O sink. It is somewhat interesting to note that the some of the highest concentrations were observed in Indiana/Ohio, where I conducted much of my dissertation research on aquatic N2O emissions.

```{r}
ggplot() +
  geom_sf(data = ecoR, color = NA, aes(fill = WSA9_NAME)) +
  geom_sf(data = dg.sf, aes(size = dissolved.n2o, color = n2o.src.snk),
          show.legend = "point") +
  ggtitle("Dissolved N2O (nM)") +
  scale_color_manual(values = c("white", "black"), name = "source/sink") +
  scale_size(name = "dissolved N2O (nM)",
             range = c(0.1, 10), # custom size range
             breaks = c(1, 10, 25, 50, 100)) # custom breaks
```

We can also look at population estimates of N2O saturation ratio by ecoregion.  Wow, the mean N2O saturation is less than one for all ecoregions and the nation!  95% CI overlap one in only 3 levels of the stratification.  This is pretty big news. 
```{r}

# extract saturation indicator, order by n2o
pop.eco.sat <- pop.eco %>%
  filter(grepl("sat", indicator, ignore.case = TRUE)) %>%
  mutate(subpopulation = factor(subpopulation, 
                             levels = filter(pop.eco, indicator == "N2O_SAT_RATIO", subpopulation != "National") %>%
                               arrange(estimate) %>%
                               select(subpopulation) %>%
                               pull() %>%
                               c("National", .)))

# look at dissolved concentration indicator
ggplot(pop.eco.sat, aes(estimate, subpopulation)) +
  geom_point() +
  geom_errorbarh(aes(xmax = ucb95pct, xmin = lcb95pct)) +
  facet_wrap(~indicator, scales = "free_x") +
  theme(axis.title = element_blank())

```

## Lagged scatterplots
Mike, can you integrate scatterplot code here?

## Semivariograms
Variograms display the variability among pairs of points at various separation distances.  I was unable to fit a model to the dissolved gas concentration variograms using the default fitting parameters in gstat.  The variograms suggest a lack of spatial structure, but I suspect the model parameterization can be improved.

```{r, warning=FALSE, message=FALSE}
# Variogram function requires sp object
dg.sp <- as(dg.sf, Class = "Spatial")

# CO2 variogram
co2.v <- variogram(dissolved.co2~1, dg.sp)
co2.v.fit <- fit.variogram(co2.v, vgm(c("Exp", "Mat", "Sph"))) # allow to choose best model.  Not sure which model warnings are associated with.
# co2.v.fit # chose sperical
co2.v.fit <- fit.variogram(co2.v, vgm("Sph")) # refit spherical, several warnings (?)

# CH4 variogram
ch4.v <- variogram(dissolved.ch4~1, dg.sp)
ch4.v.fit <- fit.variogram(ch4.v, vgm(c("Exp", "Mat", "Sph"))) # allow to choose best model.  Not sure which model warnings are associated with.
# ch4.v.fit # chose exp
ch4.v.fit <- fit.variogram(ch4.v, vgm("Exp")) # refit exponential, no convergence

# N2O variogram
n2o.v <- variogram(dissolved.n2o~1, dg.sp)
n2o.v.fit <- fit.variogram(n2o.v, vgm(c("Exp", "Mat", "Sph"))) # allow to choose best model.  Not sure which model warnings are associated with.
# n2o.v.fit # chose exp
n2o.v.fit <- fit.variogram(n2o.v, vgm("Exp")) # refit exponential, no convergence

# Plot variograms and models (if fit)
plot(co2.v, model = co2.v.fit, main = "CO2 semivariogram")
plot(ch4.v, main = "CH4 semivariogram")
plot(n2o.v, main = "N2O semivariogram")

```
